Component response rate variation underlies the stability of highly complex finite systems

### 2020-03-26 notebook/ms.tex
Make minor edits to the notation to make everything consistent

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}} or
banks\textsuperscript{\protect\hyperlink{ref-Haldane2011}{12}}). One
element of such a matrix, \(M_{ij}\), defines how component \(j\)
affects component \(i\) in the system at a point of
equilibrium\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
Off-diagonal elements (\(i \neq j\)) therefore define interactions
between components, while diagonal elements (\(i = j\)) define component
self-regulation (e.g., carrying capacity in ecological communities).
Traditionally, off-diagonal elements are assigned non-zero values with a
probability \(C\), which are sampled from a distribution with variance
\(\sigma^{2}\); diagonal elements are set to
\(-1\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{5}}.
Local system stability is assessed using eigenanalysis on
\(\mathbf{M}\), with the system being stable if the real parts of all
eigenvalues (\(\lambda\)), and therefore the leading eigenvalue
(\(\lambda_{max}\)), are negative
(\(\Re(\lambda_{max}) < 0\))\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
In a large system (high \(S\)), eigenvalues are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{15}} within a
circle centred at \(\Re = -d\) (\(-d\) is the mean value of diagonal
elements) and \(\Im = 0\), with a radius of
\(\sigma\sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{5}}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S\), \(C\), and \(\sigma\)
increase.

@@

Equation \ref{defM} thereby serves as a null model to investigate how
variation in component response rate (\(\sigma^{2}_{\gamma}\)) affects
complex systems. In the absence of such variation
(\(\sigma^{2}_{\gamma} = 0\)), \(\boldsymbol{\gamma}\) is set to the
identity matrix (diagonal elements all equal 1) and
\(\mathbf{M} = \mathbf{A}\). Under these conditions, eigenvalues of
\(\mathbf{M}\) are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{15}} in a
circle centred at \((-d, 0)\) with a radius of
\(\sigma \sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1}}
(Fig. 1a).

@@

Given \(\sigma^{2}_\gamma = 0\), \(\Re(\lambda_{max})\) decreases
linearly with \(\rho\) such
that\textsuperscript{\protect\hyperlink{ref-Tang2014c}{26}},

@@

\textbf{Simulation of random \(\mathbf{M}\) across \(\mathbf{\rho}\)}

Numerical simulations of \(\mathbf{M}\) revealed that
\(\sigma^{2}_{\gamma}\) results in a nonlinear relationship between
\(\rho\) and \(\Re(\lambda_{max})\), which can sometimes increase the
stability of the system. Figure 2 shows a comparison of
\(\Re(\lambda_{max})\) across \(\rho\) values for \(\mathbf{A}\)
(\(\sigma^{2}_{\gamma} = 0\)) versus \(\mathbf{M}\)
(\(\sigma^{2}_{\gamma} = 1/3\)) given \(S = 25\), \(C = 1\), and
\(\sigma_{A} = 0.4\). For \(-0.4 \leq \rho \leq 0.7\) (shaded region of
Fig. 2), expected \(\Re(\lambda_{max})\) was lower in \(\mathbf{M}\)
than \(\mathbf{A}\). For \(\rho \geq -0.1\), the lower bound of the
range of \(\Re(\lambda_{max})\) values also decreased given
\(\sigma^{2}_{\gamma}\), resulting in negative \(\Re(\lambda_{max})\) in
\(\mathbf{M}\) for \(\rho = -0.1\) and \(\rho = 0\). Hence, across a
wide range of system correlations, variation in the response rate of
system components had a stabilising effect.

@@

\textbf{Simulation of random \(\mathbf{M}\) across \(\mathbf{S}\)}. To
investigate the effect of \(\sigma^{2}_{\gamma}\) on stability across
systems of increasing complexity, I simulated random
\(\mathbf{M = \gamma A}\) matrices at \(\sigma_{A} = 0.4\) and \(C = 1\)
across \(S = \{2, 3, ..., 49, 50\}\). One million \(\mathbf{M}\) were
simulated for each \(S\), and the stability of \(\mathbf{A}\) vesus
\(\mathbf{M}\) was assessed given \(\gamma_{i} \sim \mathcal{U}(0, 2)\)
(\(\sigma^{2}_{\gamma} = 1/3\)). For all \(S > 10\), I found that the
number of stable random systems was higher in \(\mathbf{M}\) than
\(\mathbf{A}\) (Fig. 4; see Supplementary Information for full table of
results), and that the difference between the probabilities of observing
a stable system increased with an increase in \(S\). In other words, the
potential for \(\sigma^{2}_{\gamma}\) to affect stability increased with
system complexity and was most relevant for systems on the cusp of being
too complex to be realistically stable. For the highest values of \(S\),
nearly all systems that were stable given varying \(\gamma\) would not
have been stable given \(\gamma = 1\).

I also simulated 100000 \(\mathbf{M = \gamma A}\) for three types of
random networks that are typically interpreted as modelling three types
of interspecific ecological
interactions\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2011}{27}}.
These interaction types are competitive, mutualist, and predator-prey,
as modelled by off-diagonal elements that are constrained to be
negative, positive, or paired such that if \(A_{ij} > 0\) then
\(A_{ji} < 0\),
respectively\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
(but are otherwise identical to the purely random \(\mathbf{A}\)). As
\(S\) increased, a higher number of stable \(\mathbf{M}\) relative to
\(\mathbf{A}\) was observed for competitor and predator-prey, but not
mutualist, systems. A higher number of stable systems was observed
whenever \(S > 12\) and \(S > 40\) for competitive and predator-prey
systems, respectively (note that \(\rho < 0\) for predator-prey systems,
making stability more likely overall). The stability of mutualist
systems was never affected by \(\sigma^{2}_{\gamma}\).

@@

Network structure did not strongly modulate the effect that
\(\sigma^{2}_{\gamma}\) had on stability. For comparable magnitudes of
complexity, structured networks still had a higher number of stable
\(\mathbf{M}\) than \(\mathbf{A}\). For random networks,
\(\sigma^{2}_{\gamma}\) increased stability given \(S > 10\)
(\(\sigma = 0.4\) and \(C = 1\)), and therefore complexity
\(\sigma_{A} \sqrt{SC} \gtrapprox 1.26\). This threshold of complexity,
above which more \(\mathbf{M}\) than \(\mathbf{A}\) were stable, was
comparable for small-world networks, and slightly lower for scale-free
networks (note that algorithms for generating small-world and scale-free
networks necessarily led to varying \(C\); see methods). Varying
\(\gamma\) increased stability in cascade food webs for \(S > 27\), and
therefore at a relatively low complexity magnitudes compared to random
predator-prey networks (\(S > 40\)). Overall, network structure did not
greatly change the effect that \(\sigma^{2}_{\gamma}\) had on increasing
the upper bound of complexity within which stability might reasonably be
observed.

\textbf{System feasibility given \(\mathbf{\sigma^{2}_{\gamma}}\)} For
complex systems in which individual system components represent the
density of some tangible quantity, it is relevant to consider the
feasibility of the system. Feasibilility assumes that values of all
components are positive at
equilibrium\textsuperscript{\protect\hyperlink{ref-Grilli2017}{6},\protect\hyperlink{ref-Dougoud2018}{28},\protect\hyperlink{ref-Song2018}{29}}.
This is of particular interest for ecological communities because
population density (\(N\)) cannot take negative values, meaning that
ecological systems need to be feasible for stability to be biologically
realistic\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{28}}.
While my results are intended to be general to all complex systems, and
not restricted to species networks, I have also performed a feasibility
analysis on all matrices tested for stability. I emphasise that
\(\gamma\) is not interpreted as population density in this analysis,
but instead as a fundamental property of species life history such as
expected generation time. Feasibility was unaffected by
\(\sigma^{2}_{\gamma}\) and instead occurred with a fixed probability of
\(1/2^{S}\), consistent with a recent proof by Serván et
al.\textsuperscript{\protect\hyperlink{ref-Servan2018}{30}} (see
Supplementary Information). Hence, for pure interacting species
networks, variation in component response rate (i.e., species generation
time) does not affect stability at biologically realistic species
densities.

@@

\textbf{Construction of random and structured networks}. I used the R
programming language for all numerical simulations and
analyses\textsuperscript{\protect\hyperlink{ref-Rproject}{33}}. Purely
random networks were generated by sampling off-diagonal elements from an
i.i.d \(A_{ij} \sim \mathcal{N}(0, 0.4)\) with a probability \(C\)
(unsampled elements were set to zero). Diagonal elements \(A_{ii}\) were
set to \(-1\). Elements of \(\boldsymbol{\gamma}\) were simulated i.i.d.
from a distribution with positive support (typically
\(\gamma_{i} \sim \mathcal{U}(0, 2)\)). Random \(\mathbf{A}\) matrices
with correlated elements \(A_{ij}\) and \(A_{ji}\) were built using
Cholesky decomposition. Competitor networks in which off-diagonal
elements \(A_{ij} \geq 0\) were constructed by first building a random
\(\mathbf{A}\), then flipping the sign of any elements in which
\(A_{ij} < 0\). Similarly, mutualist networks were constructed by
building a random \(\mathbf{A}\), then flipping the sign of elements
where \(A_{ij} > 0\). Predator-prey networks were constructed by first
buing a random \(\mathbf{A}\), then flipping the sign of either
\(A_{ij}\) or \(A_{ji}\) if \(A_{ij} \times A_{ji} > 0\).

Small-world networks were constructed using the method of Watts and
Strogatz\textsuperscript{\protect\hyperlink{ref-Watts1998}{16}}. First,
a regular network\textsuperscript{\protect\hyperlink{ref-Watts1998}{16}}
was created such that components were arranged in a circle. Each
component was initially set to interact with its \(k/2\) closest
neighbouring components on each side, where \(k\) was an even natural
number (e.g., for \(k = 2\) the regular network forms a ring in which
each component interacts with its two adjacent neighbours; see
Supplemental Material for examples). Each interaction between a focal
component and its neighbour was then removed and replaced with with a
probability of \(\beta\). In replacement, a new component was randomly
selected to interact with the focal component; selection was done with
equal probability among all but the focal component. The resulting
small-world network was represented by a square \(S \times S\) binary
matrix \(\mathbf{B}\) in which 1s represented interactions between
components and 0s represented the absence of an interaction. A new
random matrix \(\mathbf{J}\) was then generated with elements \(J_{ij}\)
sampled i.i.d. from \(\mathcal{N}(0, 0.4)\). To build the interaction
matrix \(\mathbf{A}\), I used element-wise multiplication
\(\mathbf{A} = \mathbf{J} \odot \mathbf{B}\), then set
\(diag(\mathbf{A}) = -1\). I set \(k = S/12\) and simulated small-world
networks across all combinations of
\(S = \{24, 48, 72, 96, 120, 144, 168\}\) and
\(\beta = \{0, 0.01, 0.1, 0.25, 1\}\).

Scale-free networks were constructed using the method of Albert and
Barabási\textsuperscript{\protect\hyperlink{ref-Albert2002}{17}}. First,
a saturated network (all components interact with each other) of size
\(m \leq S\) was created. New components were then added sequentially to
the network; each newly added component was set to interact with \(m\)
randomly selected existing components. When the system size reached
\(S\), the distribution of the number of total interactions that
components had followed a power-law
tail\textsuperscript{\protect\hyperlink{ref-Albert2002}{17}}. The
resulting network was represented by an \(S \times S\) binary matrix
\(\mathbf{G}\), where 1s and 0s represent the presence and absence of an
interaction, respectively. As with small-world networks, a random matrix
\(\mathbf{J}\) was generated, and
\(\mathbf{A} = \mathbf{J} \odot \mathbf{G}\). Diagonal were set to
\(-1\). I simulated scale-free networks across all combinations of
\(S = \{24, 48, 72, 96, 120\}\) and \(m = \{2, 3, ..., 11, 12\}\).

Cascade food webs were constructed following Solow and
Beet\textsuperscript{\protect\hyperlink{ref-Solow1998}{18}}. First, a
random matrix \(\mathbf{A}\) was generated with off-diagonal elements
sampled i.i.d so that \(A_{ij} \sim \mathcal{N}(0, 0.4)\). Each
component in the system was ranked from \(1\) to \(S\). If component
\(i\) had a higher rank than component \(j\) and \(A_{ij} < 0\), then
\(A_{ij}\) was multiplied by \(-1\). If \(i\) had a lower rank than
\(j\) and \(A_{ji} < 0\), then \(A_{ji}\) was multiplied by \(-1\). In
practice, this resulted in a matrix \(\mathbf{A}\) with negative and
positive values in the lower and upper triangles, respectively. Diagonal
elements of \(\mathbf{A}\) were set to \(-1\) and \(C = 1\). I simulated
cascade food webs across all combinations of
\(S = \{2, 3, ..., 59, 60\}\).

@@

For each \(S = \{2, 3, ..., 39, 40\}\), the genetic algorithm was run
for 100000 random \(\mathbf{M = \gamma A}\) (\(\sigma_{A} = 0.4\),
\(C = 1\)). The genetic algorithm was initialised with a population of
1000 different \(diag(\boldsymbol{\gamma})\) vectors with elements
sampled i.i.d from \(\gamma_{i} \sim \mathcal{U}(0, 2)\). Eigenanalysis
was performed on the \(\mathbf{M}\) resulting from each
\(\boldsymbol{\gamma}\), and the 20 \(diag(\boldsymbol{\gamma})\)
vectors resulting in \(\mathbf{M}\) with the lowest
\(\Re(\lambda_{max})\) each produced 50 clonal offspring with subsequent
random mutation and crossover between the resulting new generation of
1000 \(diag(\boldsymbol{\gamma})\) vectors. Mutation of each
\(\gamma_{i}\) in a \(diag(\boldsymbol{\gamma})\) vector occurred with a
probability of 0.2, resulting in a mutation effect of size
\(\mathcal{N}(0, 0.02)\) being added to generate the newly mutated
\(\gamma_{i}\) (any \(\gamma_{i}\) values that mutated below zero were
multiplied by \(-1\), and any values that mutated above 2 were set to
2). Crossover occurred between two sets of 100
\(diag(\boldsymbol{\gamma})\) vectors paired in each generation; vectors
were randomly sampled with replacement among but not within sets. Vector
pairs selected for crossover swapped all elements between and including
two \(\gamma_{i}\) randomly selected with replacement (this allowed for
reversal of vector element positions during crossover; e.g.,
\(\{\gamma_{4}, \gamma_{5}, \gamma_{6}, \gamma_{7}\} \to \{\gamma_{7}, \gamma_{6}, \gamma_{5}, \gamma_{4}\}\)
). The genetic algorithm terminated if a stable \(\mathbf{M}\) was
found, 20 generations occurred, or if the mean \(\boldsymbol{\gamma}\)
fitness increase between generations was less than 0.01 (where fitness
was defined as \(W_{\gamma} = -\Re(\lambda_{max})\) for \(\mathbf{M}\)).

@@

\textbf{Figure 1: Eigenvalue distributions of random complex systems.}
Each panel shows the real (x-axis) and imaginary (y-axis) parts of
\(S =\) 400 eigenvalues from random \(S \times S\) matrices.
(\(\textbf{a}\)) A system represented by a matrix \(\mathbf{A}\), in
which all elements are sampled from a normal distribution with
\(\mu = 0\) and \(\sigma_{A} = 1/\sqrt{S}\). Points are uniformly
distributed within the blue circle centred at the origin with a radius
of \(\sigma_{A} \sqrt{S} = 1\). (\(\textbf{b}\)) The same system as
\(\textbf{a}\) after including variation in the response rates of \(S\)
components, represented by the diagonal matrix \(\gamma\), such that
\(\mathbf{M} = \gamma\mathbf{A}\). Elements of \(\gamma\) are randomly
sampled from a uniform distribution from \(\min = 0\) to \(\max = 2\).
Eigenvalues of \(\mathbf{M}\) are then distributed non-uniformly within
the red circle centred at the origin with a radius of
\(\sqrt{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})S} \approx\) 1.14.
(\(\textbf{c}\)) A different random system \(\mathbf{A}\) constructed
from the same parameters as in \(\textbf{a}\), except with diagonal
values of \(-1\). (\(\textbf{d}\)) The same system \(\textbf{c}\) after
including variation in component response rates, sampled from
\(\mathcal{U}(0, 2)\) as in \(\textbf{b}\).

>>>>

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}} or
banks\textsuperscript{\protect\hyperlink{ref-Haldane2011}{12}}). One
element of such a matrix, \(M_{ij}\), defines how component \(j\)
affects component \(i\) in the system at a point of
equilibrium\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
Off-diagonal elements (\(i \neq j\)) therefore define interactions
between components, while diagonal elements (\(i = j\)) define component
self-regulation (e.g., carrying capacity in ecological communities).
Traditionally, off-diagonal elements are assigned non-zero values with a
probability \(C\), which are sampled from a distribution with variance
\(\sigma^{2}\); diagonal elements are set to
\(-1\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{5}}.
Local system stability is assessed using eigenanalysis on
\(\mathbf{M}\), with the system being stable if the real parts of all
eigenvalues (\(\lambda\)), and therefore the leading eigenvalue
(\(\lambda_{max}\)), are negative
(\(\Re(\lambda_{max}) < 0\))\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
In a large system (high \(S\)), eigenvalues are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{15}} within a
circle centred at \(\Re = -d\) (\(-d\) is the mean value of diagonal
elements) and \(\Im = 0\), with a radius of
\(\sigma\sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{5}}
(Fig. 1a). Local stability of randomly assembled systems therefore
becomes increasingly unlikely as \(S\), \(C\), and \(\sigma\) increase.

@@

Equation \ref{defM} thereby serves as a null model to investigate how
variation in component response rate (\(\sigma^{2}_{\gamma}\)) affects
complex systems. In the absence of such variation
(\(\sigma^{2}_{\gamma} = 0\)), \(\boldsymbol{\gamma}\) is set to the
identity matrix (diagonal elements all equal 1) and
\(\mathbf{M} = \mathbf{A}\). Under these conditions, eigenvalues of
\(\mathbf{M}\) are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{15}} in a
circle centred at \((-1, 0)\) with a radius of
\(\sigma \sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1}}
(Fig. 1a).

@@

Given \(\sigma^{2}_\gamma = 0\), \(\Re(\lambda_{max})\) increases
linearly with \(\rho\) such
that\textsuperscript{\protect\hyperlink{ref-Tang2014c}{26}},

@@

\textbf{Simulation of random \(\mathbf{M}\) across \(\mathbf{\rho}\)}.
Numerical simulations revealed that \(\sigma^{2}_{\gamma}\) results in a
nonlinear relationship between \(\rho\) and \(\Re(\lambda_{max})\),
which can sometimes increase the stability of the system. Figure 2 shows
a comparison of \(\Re(\lambda_{max})\) across \(\rho\) values for
\(\mathbf{A}\) (\(\sigma^{2}_{\gamma} = 0\)) versus \(\mathbf{M}\)
(\(\sigma^{2}_{\gamma} = 1/3\)) given \(S = 25\), \(C = 1\), and
\(\sigma_{A} = 0.4\). For \(-0.4 \leq \rho \leq 0.7\) (shaded region of
Fig. 2), expected \(\Re(\lambda_{max})\) was lower in \(\mathbf{M}\)
than \(\mathbf{A}\). For \(\rho \geq -0.1\), the lower bound of the
range of \(\Re(\lambda_{max})\) values also decreased given
\(\sigma^{2}_{\gamma}\), resulting in negative \(\Re(\lambda_{max})\) in
\(\mathbf{M}\) for \(\rho = -0.1\) and \(\rho = 0\). Hence, across a
wide range of system correlations, variation in the response rate of
system components had a stabilising effect.

@@

\textbf{Simulation of random \(\mathbf{M}\) across \(\mathbf{S}\)}. To
investigate the effect of \(\sigma^{2}_{\gamma}\) on stability across
systems of increasing complexity, I simulated random
\(\mathbf{M = \gamma A}\) matrices at \(\sigma_{A} = 0.4\) and \(C = 1\)
across \(S = \{2, 3, ..., 49, 50\}\). One million \(\mathbf{M}\) were
simulated for each \(S\), and the stability of \(\mathbf{A}\) vesus
\(\mathbf{M}\) was assessed given \(\gamma_{i} \sim \mathcal{U}(0, 2)\)
(\(\sigma^{2}_{\gamma} = 1/3\)). For all \(S > 10\), I found that the
number of stable random systems was higher in \(\mathbf{M}\) than
\(\mathbf{A}\) (Fig. 4; see Supplementary Information for full table of
results), and that the difference between the probabilities of observing
a stable system increased with an increase in \(S\). In other words, the
potential for \(\sigma^{2}_{\gamma}\) to affect stability increased with
increasing system complexity and was most relevant for systems on the
cusp of being too complex to be realistically stable. For the highest
values of \(S\), nearly all systems that were stable given varying
\(\gamma\) would not have been stable given \(\gamma = 1\).

I also simulated 100000 \(\mathbf{M}\) for three types of random
networks that are typically interpreted as modelling three types of
interspecific ecological
interactions\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2011}{27}}.
These interaction types are competitive, mutualist, and predator-prey,
as modelled by off-diagonal elements that are constrained to be
negative, positive, or paired such that if \(A_{ij} > 0\) then
\(A_{ji} < 0\),
respectively\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
(but are otherwise identical to the purely random \(\mathbf{A}\)). As
\(S\) increased, a higher number of stable \(\mathbf{M}\) relative to
\(\mathbf{A}\) was observed for competitor and predator-prey, but not
mutualist, systems. A higher number of stable systems was observed
whenever \(S > 12\) and \(S > 40\) for competitive and predator-prey
systems, respectively (note that \(\rho < 0\) for predator-prey systems,
making stability more likely overall). The stability of mutualist
systems was never affected by \(\sigma^{2}_{\gamma}\).

@@

Network structure did not strongly modulate the effect that
\(\sigma^{2}_{\gamma}\) had on stability. For comparable magnitudes of
complexity, structured networks still had a higher number of stable
\(\mathbf{M}\) than \(\mathbf{A}\). For random networks,
\(\sigma^{2}_{\gamma}\) increased stability given \(S > 10\)
(\(\sigma_{A} = 0.4\) and \(C = 1\)), and therefore complexity
\(\sigma_{A} \sqrt{SC} \gtrapprox 1.26\). This threshold of complexity,
above which more \(\mathbf{M}\) than \(\mathbf{A}\) were stable, was
comparable for small-world networks, and slightly lower for scale-free
networks (note that algorithms for generating small-world and scale-free
networks necessarily led to varying \(C\); see methods). Varying
\(\gamma\) increased stability in cascade food webs for \(S > 27\), and
therefore at a relatively low complexity magnitudes compared to random
predator-prey networks (\(S > 40\)). Overall, network structure did not
greatly change the effect that \(\sigma^{2}_{\gamma}\) had on increasing
the upper bound of complexity within which stability might reasonably be
observed.

\textbf{System feasibility given \(\mathbf{\sigma^{2}_{\gamma}}\)} For
complex systems in which individual system components represent the
density of some tangible quantity, it is relevant to consider the
feasibility of the system. Feasibilility assumes that values of all
components are positive at
equilibrium\textsuperscript{\protect\hyperlink{ref-Grilli2017}{6},\protect\hyperlink{ref-Dougoud2018}{28},\protect\hyperlink{ref-Song2018}{29}}.
This is of particular interest for ecological communities because
population density (\(n\)) cannot take negative values, meaning that
ecological systems need to be feasible for stability to be biologically
realistic\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{28}}.
While my results are intended to be general to all complex systems, and
not restricted to species networks, I have also performed a feasibility
analysis on all matrices tested for stability. I emphasise that
\(\gamma\) is not interpreted as population density in this analysis,
but instead as a fundamental property of species life history such as
expected generation time. Feasibility was unaffected by
\(\sigma^{2}_{\gamma}\) and instead occurred with a fixed probability of
\(1/2^{S}\), consistent with a recent proof by Serván et
al.\textsuperscript{\protect\hyperlink{ref-Servan2018}{30}} (see
Supplementary Information). Hence, for pure interacting species
networks, variation in component response rate (i.e., species generation
time) does not affect stability at biologically realistic species
densities.

@@

\textbf{Construction of random and structured networks}. I used the R
programming language for all numerical simulations and
analyses\textsuperscript{\protect\hyperlink{ref-Rproject}{33}}. Purely
random networks were generated by sampling off-diagonal elements from an
i.i.d \(A_{ij} \sim \mathcal{N}(0, 0.4^{2})\) with a probability \(C\)
(unsampled elements were set to zero). Diagonal elements \(A_{ii}\) were
set to \(-1\). Elements of \(\boldsymbol{\gamma}\) were simulated i.i.d.
from a distribution with positive support (typically
\(\gamma_{i} \sim \mathcal{U}(0, 2)\)). Random \(\mathbf{A}\) matrices
with correlated elements \(A_{ij}\) and \(A_{ji}\) were built using
Cholesky decomposition. Competitor networks in which off-diagonal
elements \(A_{ij} \leq 0\) were constructed by first building a random
\(\mathbf{A}\), then flipping the sign of any elements in which
\(A_{ij} > 0\). Similarly, mutualist networks were constructed by
building a random \(\mathbf{A}\), then flipping the sign of elements
where \(A_{ij} < 0\). Predator-prey networks were constructed by first
building a random \(\mathbf{A}\), then flipping the sign of either
\(A_{ij}\) or \(A_{ji}\) if \(A_{ij} \times A_{ji} > 0\).

Small-world networks were constructed using the method of Watts and
Strogatz\textsuperscript{\protect\hyperlink{ref-Watts1998}{16}}. First,
a regular network\textsuperscript{\protect\hyperlink{ref-Watts1998}{16}}
was created such that components were arranged in a circle. Each
component was initially set to interact with its \(k/2\) closest
neighbouring components on each side, where \(k\) was an even natural
number (e.g., for \(k = 2\) the regular network forms a ring in which
each component interacts with its two adjacent neighbours; see
Supplemental Material for examples). Each interaction between a focal
component and its neighbour was then removed and replaced with with a
probability of \(\beta\). In replacement, a new component was randomly
selected to interact with the focal component; selection was done with
equal probability among all but the focal component. The resulting
small-world network was represented by a square \(S \times S\) binary
matrix \(\mathbf{B}\) in which 1s represented interactions between
components and 0s represented the absence of an interaction. A new
random matrix \(\mathbf{J}\) was then generated with elements \(J_{ij}\)
sampled i.i.d. from \(\mathcal{N}(0, 0.4^{2})\). To build the
interaction matrix \(\mathbf{A}\), I used element-wise multiplication
\(\mathbf{A} = \mathbf{J} \odot \mathbf{B}\), then set
\(diag(\mathbf{A}) = -1\). I set \(k = S/12\) and simulated small-world
networks across all combinations of
\(S = \{24, 48, 72, 96, 120, 144, 168\}\) and
\(\beta = \{0, 0.01, 0.1, 0.25, 1\}\).

Scale-free networks were constructed using the method of Albert and
Barabási\textsuperscript{\protect\hyperlink{ref-Albert2002}{17}}. First,
a saturated network (all components interact with each other) of size
\(m \leq S\) was created. New components were then added sequentially to
the network; each newly added component was set to interact with \(m\)
randomly selected existing components. When the system size reached
\(S\), the distribution of the number of total interactions that
components had followed a power-law
tail\textsuperscript{\protect\hyperlink{ref-Albert2002}{17}}. The
resulting network was represented by an \(S \times S\) binary matrix
\(\mathbf{G}\), where 1s and 0s represent the presence and absence of an
interaction, respectively. As with small-world networks, a random matrix
\(\mathbf{J}\) was generated, and
\(\mathbf{A} = \mathbf{J} \odot \mathbf{G}\). Diagonal elements were set
to \(-1\). I simulated scale-free networks across all combinations of
\(S = \{24, 48, 72, 96, 120\}\) and \(m = \{2, 3, ..., 11, 12\}\).

Cascade food webs were constructed following Solow and
Beet\textsuperscript{\protect\hyperlink{ref-Solow1998}{18}}. First, a
random matrix \(\mathbf{A}\) was generated with off-diagonal elements
sampled i.i.d so that \(A_{ij} \sim \mathcal{N}(0, 0.4^2)\). Each
component in the system was ranked from \(1\) to \(S\). If component
\(i\) had a higher rank than component \(j\) and \(A_{ij} < 0\), then
\(A_{ij}\) was multiplied by \(-1\). If \(i\) had a lower rank than
\(j\) and \(A_{ji} < 0\), then \(A_{ji}\) was multiplied by \(-1\). In
practice, this resulted in a matrix \(\mathbf{A}\) with negative and
positive values in the lower and upper triangles, respectively. Diagonal
elements of \(\mathbf{A}\) were set to \(-1\) and \(C = 1\). I simulated
cascade food webs for \(S = \{2, 3, ..., 59, 60\}\).

@@

For each \(S = \{2, 3, ..., 39, 40\}\), the genetic algorithm was run
for 100000 random \(\mathbf{M = \gamma A}\) (\(\sigma_{A} = 0.4\),
\(C = 1\)). The genetic algorithm was initialised with a population of
1000 different \(diag(\boldsymbol{\gamma})\) vectors with elements
sampled i.i.d from \(\gamma_{i} \sim \mathcal{U}(0, 2)\). Eigenanalysis
was performed on the \(\mathbf{M}\) resulting from each
\(\boldsymbol{\gamma}\), and the 20 \(diag(\boldsymbol{\gamma})\)
vectors resulting in \(\mathbf{M}\) with the lowest
\(\Re(\lambda_{max})\) each produced 50 clonal offspring with subsequent
random mutation and crossover between the resulting new generation of
1000 \(diag(\boldsymbol{\gamma})\) vectors. Mutation of each
\(\gamma_{i}\) in a \(diag(\boldsymbol{\gamma})\) vector occurred with a
probability of 0.2, resulting in a mutation effect of size
\(\mathcal{N}(0, 0.02^{2})\) being added to generate the newly mutated
\(\gamma_{i}\) (any \(\gamma_{i}\) values that mutated below zero were
multiplied by \(-1\), and any values that mutated above 2 were set to
2). Crossover occurred between two sets of 100
\(diag(\boldsymbol{\gamma})\) vectors paired in each generation; vectors
were randomly sampled with replacement among but not within sets. Vector
pairs selected for crossover swapped all elements between and including
two \(\gamma_{i}\) randomly selected with replacement (this allowed for
reversal of vector element positions during crossover; e.g.,
\(\{\gamma_{4}, \gamma_{5}, \gamma_{6}, \gamma_{7}\} \to \{\gamma_{7}, \gamma_{6}, \gamma_{5}, \gamma_{4}\}\)
). The genetic algorithm terminated if a stable \(\mathbf{M}\) was
found, 20 generations occurred, or if the mean \(\boldsymbol{\gamma}\)
fitness increase between generations was less than 0.01 (where fitness
was defined as \(W_{\gamma} = -\Re(\lambda_{max})\) for \(\mathbf{M}\)).

@@

\textbf{Figure 1: Eigenvalue distributions of random complex systems.}
Each panel shows the real (x-axis) and imaginary (y-axis) parts of
\(S =\) 400 eigenvalues from random \(S \times S\) matrices.
(\(\textbf{a}\)) A system represented by a matrix \(\mathbf{A}\), in
which all elements are sampled from a normal distribution with
\(\mu = 0\) and \(\sigma_{A} = 1/\sqrt{S}\). Points are uniformly
distributed within the blue circle centred at the origin with a radius
of \(\sigma_{A} \sqrt{S} = 1\). (\(\textbf{b}\)) The same system as
\(\textbf{a}\) after including variation in the response rates of \(S\)
components, represented by the diagonal matrix \(\gamma\), such that
\(\mathbf{M} = \gamma\mathbf{A}\). Elements of \(\gamma\) are randomly
sampled from a uniform distribution from \(\min = 0\) to \(\max = 2\).
Eigenvalues of \(\mathbf{M}\) are then distributed non-uniformly within
the red circle centred at the origin with a radius of
\(\sqrt{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})S} \approx\) 1.19.
(\(\textbf{c}\)) A different random system \(\mathbf{A}\) constructed
from the same parameters as in \(\textbf{a}\), except with diagonal
element values of \(-1\). (\(\textbf{d}\)) The same system
\(\textbf{c}\) after including variation in component response rates,
sampled from \(\mathcal{U}(0, 2)\) as in \(\textbf{b}\).
